### 2018-04-16 contents/abstract.tex
Updates from ShareLaTeX

Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of TN revealed diffusivity changes in the cistern CNV. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of TN in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we discover the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework. Finally, in Study IV, we present an end-to-end machine-learning TN classification study that auto-discovers diffusivity disruptions in the cistern/REZ CNV and thalamocortico S1 pathway. 
In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in deeper white matter structures, and pin-points the exact location of the diffusivity disruption at millimetre level.

>>>>

Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework. Finally, in Study IV, we deploy end-to-end machine-learning TN classification to auto-discovers diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway. 
In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.

### 2018-04-16 contents/acknowledgements.tex
Updates from ShareLaTeX

First I thank my parents. They have sacrificed their careers and endured much hardship by immigrating to Canada, to provide me with a better future. All my achievements are only possible after I stand on their shoulders.

I met my amazing wife Jidan Zhong in 2012, at Human Brain Mapping conference in Beijing. Little did I know that I had found my kindred spirit. Without her continue to support and persistent encouragement, this thesis would not be possible.

Importantly, I thank my mentor Dr. Mojgan Hodaie. She took a chance on me when I had zero experience in neuroscience, and graciously supported me while I was having a hard time being accepted into the graduate program. I also thank her incredible patience and support for during my thesis writing. Dr. Hodaie is the most accomplished and hard-working person I have ever known. She is a world-class neurosurgeon, while at the same time balances her time between graduate mentoring, grant writing, international neurosurgical education, and at the same time multiple projects in scientific research. Every time I feel dejected from the workload, I always remind myself that Moji is doing more and that I have no excuses. It is truly my fortune to have her as my PI, and words cannot express how thankful I am.

I would like to thank my graduate committee members Dr. Karen Davis and Dr. David Mikulis. Whose extensible experience and insights were invaluable for guiding me through this process. Dr. Davis' impressive scientific acumen, and Dr. Mikulis' enthusiasm will forever stay with me.

I would also like to thank the past and present graduate and post-graduate members of the Hodaie Lab, whom I have shared an unforgettable period of my life: Danielle DeSouza, Allie Rosen, David Hayes, Brendan Behan, Matt Walker and Peter Hung. I also thank Jessica Quan, who introduced me to Dr. Hodaie and spark my journey in Neuroscience.

I was fortunate and thankful to receive support from Multiple Sclerosis Society of Canada, James F. Crother's Family Fellowship in peripheral nerve damage, and Unilever/Lipton graduate fellowships.

I thank all the research assistants over the years that helped me to perform the thankless job of download and organize patient data: Philip Yee, Udi Blankstein, Tessa Burkholder, Kevin Liang, and Erika Wharton-Shukster.

>>>>

First I thank my parents. They have sacrificed their careers and endured much hardships by immigrating to Canada to provide me with a better future. All my achievements are only possible after I stand on their shoulders.

I met my amazing wife Jidan Zhong in 2012, at the Human Brain Mapping conference in Beijing. Little did I know that I had found my kindred spirit. Without her continue to support and persistent encouragement, this thesis would not be possible.

Importantly, I thank my mentor Dr. Mojgan Hodaie. She took a chance on me when I had zero experience in neuroscience, and graciously supported me while I was having a hard time being accepted into the graduate program. I also thank her incredible patience and support during my thesis writing. Dr. Hodaie is the most accomplished and hard-working person I have ever known. She is a world-class neurosurgeon, while at the same time balances her time between graduate mentoring, grant writing, international neurosurgical education, and multiple projects in scientific research. Every time I feel dejected from the workload, I always remind myself that Moji is doing more and that I have no excuses. It is truly my fortune to have her as my supervisor, and words cannot express how thankful I am.

I would like to thank my graduate committee members Dr. Karen Davis and Dr. David Mikulis, whose extensive experience and insights were invaluable for guiding me through this process. Dr. Davis' impressive scientific acumen, and Dr. Mikulis' enthusiasm will forever stay with me.

I would also like to thank the past and present graduate and post-graduate members of the Hodaie Lab, with whom I have shared an unforgettable period of my life: Danielle DeSouza, Allie Rosen, David Hayes, Brendan Behan, Matt Walker and Peter Hung. I also thank Jessica Quan, who introduced me to Dr. Hodaie and sparked my journey in Neuroscience.

I was fortunate and thankful to receive support from the Multiple Sclerosis Society of Canada, James F. Crother's Family Fellowship in peripheral nerve damage, and Unilever/Lipton graduate fellowships.

I thank all the research assistants over the years that helped me to perform the thankless job of downloading and organizing data: Philip Yee, Udi Blankstein, Tessa Burkholder, Kevin Liang, and Erika Wharton-Shukster.

### 2018-04-16 contents/discussions.tex
Updates from ShareLaTeX

The contribution of this thesis are as follows:

@@

In sum, this thesis provides novel evidence on the diffusivity tissue disruptions of the CN V in TN, and that there is evidence of bilateral CN V tissue changes in TN patients when compared to controls. The examination was also extended to the thalamocortical S1 pathway and have demonstrated localized white matter changes near the white-grey matter boundary. These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies as a whole. These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter, and that the methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

>>>>

The contributions of this thesis are as follows:

@@

In sum, this thesis provides novel evidence for diffusivity tissue disruptions of the CN V in TN, and bilateral CN V tissue abnormalities in TN patients when compared to controls. The study of the trigeminal pontine decussation (TPT) and thalamocortical S1 pathway demonstrated white matter diffusivity changes at all levels of the trigeminal sensory pathway. These examinations on the group level were made possible by the development of the SAGIT merged tractography framework, that can be generally applied to group-based white matter studies. These results demonstrate that TN involves wide-spread changes at all levels of CNS white matter. The methods developed in this thesis may lead to improvements in white matter disease detection and diagnosis.

### 2018-04-16 contents/hypotheses.tex
Updates from ShareLaTeX

The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease. This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls. With an increasing number of subjects, single subject analysis becomes cumbersome. Critical methodological advances are necessary for the feasibility of the general aim. These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy. Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.  The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, to identify specific diffusivity pattern that differentiates between TN and non-TN. The roadmap and rationale for the specific studies are described below.

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate pathophysiologically to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

\subsection{Main Aim} 
Establish the feasibility of multi-tensor diffusion tractography in delineating the cisternal, root entry zone and brainstem segments of CN V and how these changes correlate with pain in MS-TN.

\subsection{Specific Aim}
\begin{itemize}
    \item Use multi-tensor tractography to delineate CN V into the the pontine brainstem, in order to quantify tissue diffusivity measurements in targeted sub-regions.
    \item Differentiate TN and MS-TN based on the obtained CN V diffusivity measurements.
\end{itemize}

@@

\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
The construction of an automated group tractography software framework depends on the establishment of the feasiblity of CN V brainstem delineation using multi-tensor tractography, and the optimal DWI-T1 co-registration strategy. The software aims to automate the tractography process on a large-scale, and also aims to streamline research workflow by providing region-of-interest, and tractography parameter management. It also provides the ability to report the tractography output at a group level both visually, as well as quantifying the output by an original scalar score called Normalized Overlap Score (NOS) score.

\subsection{Main Aim}
The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population. The tasks are performed  at a scale that would be prohibitive with manual methods. The comparisons determine the best tractography methods to use for different types of neuroanatomy.

\subsection{Specific Aim}
\begin{itemize}
    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD in 42 subjects. 
    \item Use the above techniques to study a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
    \item Validate NOS scores by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item DTI is known to not able to delineate cross-fibers regions. If the group tractography works consistently, then its output should reflect such known algorithmic limitations, and DTI group tracts should not be able to perform well on more complex anatomies. The grouped tracts should however show consistent regional anatomical overlap across individuals.
    \item By merging the tracts of many individuals, the tractography delinations can complement each other, and this may improve the anatomical fidelity over individual cases.
    \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain. For examples, deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
    \item NOS score should correlate with expert ratings, and show improved rating consistency over human rater variability.
\end{itemize}

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The developed methodology is applied to the study of the trigeminal sensory pathway. The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study. The study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. With the increased datapoints, more detailed along-the-tract analysis can be carried out on each segment. Following this, we use machine learning algorithms to auto-determine relevant segments that maximally differentiates TN from controls.

\subsection{Main Aim}
The aim of this project is the reconstruction of the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and examination using Gaussian Process classification.

\subsection{Specific Aim}
\begin{itemize}
    \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamic and  thalamocortical S1 segment. 
    \item At each segment, we will look for microstructural features to distinguish the affected vs. unaffected sides.
\end{itemize}

>>>>

The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease. This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls.

With an increasing number of subjects, single subject analysis becomes cumbersome. Critical methodological advances are necessary for the feasibility of the general aim. These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy.

Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.  The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, in order to identify specific diffusivity pattern that differentiates between TN and non-TN. The roadmap and rationale for the specific studies are described below.

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

\subsection{Main Aim} 
Establish the feasibility of multi-tensor diffusion tractography to delineating the cisternal, root entry zone and brainstem segments of CN V and how diffusivity changes in these sub-regions correlate with pain in MS-TN.

\subsection{Specific Aim}
\begin{itemize}
    \item Use multi-tensor tractography to delineate CN V into the the pontine brainstem, in order to quantify tissue diffusivity measurements in targeted sub-regions.
    \item Differentiate TN from MS-TN based on CN V diffusivity measurements.
\end{itemize}

@@

\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
The construction of an automated group tractography software framework depends on CN V brainstem delineation using multi-tensor tractography and the optimal DWI-T1 co-registration strategy. The aim is to create and validate a software framework that automates the tractography process on a large-scale, and also streamline research workflow by providing region-of-interest, and tractography parameter management. The software also provides the ability to report the tractography output at a group level both visually, as well as quantifying the consistency of group tractography output by developing a rating score called Normalized Overlap Score (NOS) score.

\subsection{Main Aim}
The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population. The validations are performed  at a scale that would be prohibitive with manual methods. The comparisons determine the best tractography methods to use for different types of neuroanatomy.

\subsection{Specific Aim}
\begin{itemize}
    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD. 
    \item Use the above techniques to delineate a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
    \item Validate normalized overlapping scores (NOS) by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item DTI is known to not able to delineate cross-fibers regions. If the group tractography works consistently, then its output should reflect such known algorithmic limitations, and DTI group tracts should not be able to perform well on more complex anatomies. The grouped tracts should however show consistent regional anatomical overlap across individuals.
    \item By merging the tracts of many individuals, the tractography delineations can complement each other, and this may improve the anatomical fidelity over individual cases.
    \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain. For examples, deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
    \item NOS score should correlate with expert ratings, and show improved rating consistency over human rater variability.
\end{itemize}

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The developed methodology is applied to the study of the trigeminal sensory pathway. The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study. This study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. Along-the-tract analysis are carried out by dividing each segment into 30 subdivisions. Following this, we use machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.

\subsection{Main Aim}
The aim of this project is to reconstruct the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and to examination these pathways using Gaussian Process classification to determine the subdivisions along each pathways that maximally differentiates TN from controls.

\subsection{Specific Aim}
\begin{itemize}
    \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamic and  thalamocortical S1 segment. 
    \item At each segment, identify microstructural features to distinguish the affected vs. unaffected sides.
\end{itemize}

### 2018-04-16 contents/paper1.tex
Updates from ShareLaTeX

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure2.png}
\caption{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views. The reconstructed tracts are overlaid onto the axial DTI/anatomical images. The colors of the underlying scan are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. The colors of the tractography fibers represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner. Panel A shows that SDT tractography delineated the cerebellar peduncle fibers, but cannot distinguish the brainstem trigeminal fibers (yellow arrows). Red arrows denote the starting point of tract generation for both SDT and MTT methods. Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei (yellow arrows).}
\centering
\label{fig:MSfigure2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure3.png}
\caption{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method. Images are shown in superior axial view. Panel A shows the MS-TN tractography delineations. There is focal decrease in CN V FA in the region of the MS lesion (yellow arrow). Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment. Please refer to Figure 2 for explanations of color scales and figure annotations.}
\centering
\label{fig:MSfigure3}
\end{figure}

>>>>

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure2.png}
\caption{Reconstruction of the trigeminal nerves at the level of pons from a single healthy subject in superior axial views. The reconstructed tracts are overlaid onto the axial DTI/anatomical images. The colors of the underlying scan are rendered as color-by-orientation, where by convention red represents left–right, greens represents anterior–posterior and blue represents inferior–superior orientations. The colors of the tractography fibers represent the spectrum values of FA (0 to 1), as shown by the legend at the top right corner. Panel A shows that SDT tractography delineated the cerebellar peduncle fibers, but cannot distinguish the brainstem trigeminal fibers (yellow arrows). Red arrows denote the starting point of tract generation for both SDT and MTT methods. Panel B shows that MTT tractography visualized CN V as it coursed through the brainstem towards the trigeminal main sensory nuclei.}
\centering
\label{fig:MSfigure2}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure3.png}
\caption{Comparison of tractography delineation between affected and unaffected CN V nerve in the region of the trigeminal main sensory nucleus using the MTT method. Images are shown in superior axial view. Panel A shows the MS-TN tractography delineations. There is focal decrease in CN V FA in the region of the MS lesion (yellow arrow). Panel B shows the TN tractography delineations, where there is no such FA decrease in the brainstem segment. Please refer to Figure \protect\ref{fig:MSfigure2} for explanations of color scales and figure annotations.}
\centering
\label{fig:MSfigure3}
\end{figure}
